### 2021-12-08 text/outline/draft.tex
complete draft

@@

We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data. We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d. We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (result replicated across nearly all pairs; average relative difference $0.003$). This suggested that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets. Next, we observed that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable on the genes \textit{LRIF1} and \textit{NDUFA2}.

We generalized the above analysis to the entire set of positive control pairs. First, for each pair, we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d). Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC). As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable. Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate). GLM-EIV exhibited significantly higher CI coverage rates than thresholded regression as the data became increasingly contaminated.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
	\caption{\textbf{Data analysis}. \textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs. \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination. Grey bands, 95\% CIs for the target of inference outputted by the methods. \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination. Colored bands, pointwise 95\% CIs for population quantities. Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}
	\label{fig:real_data}
\end{figure}

@@

% We see the practical use of this method as follows. First, deploy GLM-EIV to some random subset of pairs to assess the extent of background contamination. If the background contamination sufficiently high, apply GLM-EIV to analyze the data directly. On the other hand, if the background contamination is low, the thresholding method is an appealing method due to its speed and stability. A challenge, as we have shown, is that the selected threshold exerts a substantial impact on the results. GLM-EIV approximates the Bayes-optimal decision boundary, which, as we showed in an idealized Gaussian setting, minimizes attenuation bias when the gRNA count distribution is sufficiently-well separated. Therefore, in problem settings where thresholding is tenable, GLM-EIV provides a statistically principled, scientifically reproducible, empirically effective threshold selection procedure.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{../../figures/dec_tree/dec_tree_crop}
\caption{\textbf{Use of GLM-EIV in practice}. The decision tree above illustrates how we anticipate GLM-EIV could be used in practice. First, apply GLM-EIV to a set of randomly-sampled gene-perturbation pairs to assess background contamination level (positive control pairs work best for this purpose). If GLM-EIV indicates that background contamination is high (e.g., $\exp(\beta^g_1) \lessapprox $ 10), apply GLM-EIV to analyze the entire dataset; otherwise, approximate the Bayes-optimal decision boundary using the fitted GLM-EIV models. Next, apply a thresholding method (e.g., SCEPTRE or thresholded negative binomial regression) to analyze the data, setting the threshold to the estimated Bayes-optimal decision boundary.}
\end{figure}

% To our knowledge, this is the first statistical work to study single-cell CRISPR screens. One of our objectives is to bring to the attention of the statistical community an emerging class of functional genomics assays that likely will exert a major impact on biology research. Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a set of programming tools for large-scale single-cell analysis that we view as powerful and user-friendly.

>>>>

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
	\caption{\textbf{Data analysis}. \textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs. \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination. Grey bands, 95\% CIs for the target of inference outputted by the methods. \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination. Colored bands, pointwise 95\% CIs for population quantities. Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
\end{figure}

@@

We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data. We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d. We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We concluded that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets. We found that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.

We generalized the above analysis to the entire set of positive control pairs. First, for each pair we computed its ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d). Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC). As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable. Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate). GLM-EIV exhibited a significantly higher CI coverage rate than thresholded regression as the data became increasingly contaminated.

@@

In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model (when the latent predictor is binary) to arbitrary exponential distributions and link functions, facilitating the analysis of single-cell CRISPR screen data. We demonstrated through theoretical analyses, simulation studies, and real data analyses that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings. On the other hand, in low background contamination settings, GLM-EIV provides a statistically principled and empirically effective procedure for selecting a threshold for use in a thresholding method (e.g., thresholded negative binomial regression, as studied in this work, or even SCEPTRE \cite{Barry2020}). GLM-EIV thereby neutralizes a tuning parameter that exerts a significant impact on the results. Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.

To our knowledge this is the first paper to examine single-cell CRISPR screens from a rigorous statistical perspective. We hope that this work can serve as an introduction to the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}. Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, an (in our view) tightly-integrated, user-friendly, and powerful set of computational tools for out-of-memory and massively distributed single-cell analysis.

We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data. For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infenction'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work). We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility data. A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}. We do not directly observe chromatin state; instead, we observe transposase-tagged DNA fragments, counts that serve as proxies for whether a given region of chromatin is open or closed. GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{../../figures/dec_tree/dec_tree_crop}
\caption{\textbf{Use of GLM-EIV in practice}. The decision tree above illustrates how we anticipate GLM-EIV could be used in practice. First, apply GLM-EIV to a set of randomly-sampled gene-perturbation pairs to assess background contamination level (positive control pairs work best for this purpose). If GLM-EIV indicates that background contamination is high (e.g., $\exp(\beta^g_1) \lessapprox $ 10), apply GLM-EIV to analyze the entire dataset; otherwise, approximate the Bayes-optimal decision boundary using the fitted GLM-EIV models. Next, apply a thresholding method (e.g., SCEPTRE or thresholded negative binomial regression) to analyze the data, setting the threshold to the estimated Bayes-optimal decision boundary.}\label{fig:dec_tree}
\end{figure}
